---
title: "Visualising numeric data"
subtitle: "Introduction to Global Health Data Science"
date: "2026-08-26"
format:
  revealjs:
    output-file: w01-03-viznum-slides.html
    theme: slides.scss
    transition: fade
    slide-number: true
    pdf-separate-fragments: true
    toc: false
    auto-stretch: false
  html:
    code-link: true
  filters: 
    - ../remove-fmt-skip.lua
  editor_options: 
    chunk_output_type: console
---

```{r, child="setup.Rmd", echo=FALSE}
```


```{r}
#| label: packages
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(openintro)

life <- readr::read_csv("../data/lifeexpectancy_infant.csv")
regmapping <- readr::read_csv("../data/loc.csv")

lifespan <- life %>%
  left_join(regmapping, by = "location")

lifespan2023=lifespan %>%
  filter(year=='2023') %>%
  select(-year)

lifeexpwide <- lifespan %>%
  pivot_wider(
    names_from = sex,
    values_from = lifeexp
  )

lifeexpwide2023 <- lifespan %>%
  filter(year == 2023) %>%
  pivot_wider(
    names_from = sex,
    values_from = lifeexp
  )
```

# Terminology

---

## Number of variables involved

- Univariate data analysis - distribution of single variable
- Bivariate data analysis - relationship between two variables
- Multivariable/multivariate data analysis - relationship between many variables at once, sometimes focusing on the relationship between two while conditioning for others. (Often we reserve *multivariate* for multiple outcomes, and *multivariable* for multiple predictors.)

---

## Types of variables

- **Numerical variables** can be classified as **continuous** or **discrete** based on whether or not the variable can take on an infinite number of values or only a finite number of distinct values (e.g., counts), respectively.
- If the variable is **categorical**, we can determine if it is **ordinal** based on whether or not the levels have a natural ordering.

---


# Data

---


## Data: Life Expectancy

:::: columns

::: {.column width="40%"}

- We focus again on the IHME data on estimated life expectancy for a variety of countries and locations worldwide in the year 2023.

:::

::: {.column width="60%"}

```{r}
#| output-lines: 18

glimpse(lifespan2023)
```

:::

::::

---

## Selected variables

<br>

<div style="font-size:0.85em;">

| Variable | Description |
|----------|-------------|
| `location` | Country or area name |
| `worldbankregion` | World region, as classified by the World Bank |
| `pop` | Estimated 2023 population of the location |
| `sex` | Binary sex as reported by the country |
| `lifeexp` | Life expectancy from infancy |

</div>

---

## Variable types

<br>

<div style="font-size:0.85em;">

| Variable | Type |
|----------|------|
| `location` | Categorical, not ordinal |
| `worldbankregion` | Categorical, not ordinal |
| `pop` | Numerical, discrete |
| `sex` | Categorical, not ordinal |
| `lifeexp` | Numerical, continuous |

</div>

In the full data (not subset to 2023), `year` is an additional numerical, discrete variable.

---

# Visualizing numerical data

---

## Describing shapes of numerical distributions

- **Shape**
  - Skewness: right-skewed, left-skewed, symmetric (skew is to the side of the longer tail)
  - Modality: unimodal, bimodal, multimodal, uniform
- **Center:** mean (`mean`), median (`median`), mode
- **Spread:** range (`range`), standard deviation (`sd`), variance (square of `sd`), interquartile range (`IQR`)
- **Unusual observations**

---

## Measures of center

Consider a random variable $x_i$ that is the estimated life expectancy for country $i$, $i = 1, \ldots, n$, among $n$ countries.

- **Mean:** estimated as

$$ \bar{x} = \frac{\sum_{i=1}^{n} x_i}{n}$$

---

## Measures of center

- **Median:** middle number (50th percentile), $x_{(n+1)/2}$
  - For $n$ odd, the median is the middle number.
  - For $n$ even, the median is the mean of the two middle numbers.
  - Just remember: it splits the ordered data in half.

- **Mode:** most frequent value in the data set

---

## Measures of spread

- variance: average squared distance from mean

- standard deviation (sd): square root of variance (on same scale as data). The sample variance $s^2$ is estimated in a single sample as $$\frac{\sum_{i=1}^n (x_i-\overline{x})^2}{n-1}$$ 

  - we think of 1 sd as a sizeable difference. For example, height of college students in the US has a sd $\approx$ 3 in men and $\approx$ 2.5 in women

---

## Measures of spread

- range: difference between highest and lowest values, e.g. $x_{(n)}-x_{(1)}$

- interquartile range (IQR): difference between 75th and 25th %iles

  - we think of this as pretty big too
  - 75th %ile of height among male college students in the US is 5'11", and the 25th %ile is 5'7", so the IQR is 4"
  - for female students, the 75th %ile is 5'5.6", 25th %ile is 5'1.6", and IQR is also 4"

---

# Visualizing relationships among numerical variables

---

## Scatterplot

Previously we viewed a scatterplot showing the relationship between life expectancy of females and males in each location.

:::: columns

::: {.column width="60%"}

```{r}
#| eval: false

ggplot(
  lifeexpwide2023,
  aes(
    x = Female,
    y = Male
  )
) +
  geom_point() +
  labs(
    title = "Life expectancy",
    subtitle = "2023",
    x = "Female life expectancy",
    y = "Male life expectancy"
  )
```

:::

::: {.column width="40%"}

```{r}
#| echo: false
#| warning: false
#| out-width: 100%

ggplot(
  lifeexpwide2023,
  aes(
    x = Female,
    y = Male
  )
) +
  geom_point() +
  labs(
    title = "Life expectancy",
    subtitle = "2023",
    x = "Female life expectancy",
    y = "Male life expectancy"
  )
```

:::

::::

---

# Histogram

---

## Histogram

:::: columns

::: {.column width="60%"}

```{r}
#| eval: false

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(x = lifeexp)
  ) +
  geom_histogram()
```

:::

::: {.column width="40%"}

```{r}
#| echo: false
#| message: false
#| out-width: 100%

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(x = lifeexp)
  ) +
  geom_histogram()
```

:::

::::

---

## Histograms and binwidth: `binwidth = 0.5`

:::: columns

::: {.column width="60%"}

The bin width is the width of the interval used for grouping the data. It is selected automatically by R, but we can alter it. For example, a bin width of $\frac{1}{2}$ plots a bar for each 6 month life expectancy group, while a bin width of 10 plots a bar for groups of 10 years.

```{r}
#| eval: false

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(x = lifeexp)
  ) +
  geom_histogram(binwidth = 0.5)
```

:::

::: {.column width="40%"}

```{r}
#| echo: false
#| warning: false
#| out-width: 100%

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(x = lifeexp)
  ) +
  geom_histogram(binwidth = 0.5)
```

:::

::::

---

## Histograms and binwidth: `binwidth = 3`

:::: columns

::: {.column width="60%"}

```{r}
#| eval: false

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(x = lifeexp)
  ) +
  geom_histogram(binwidth = 3)
```

:::

::: {.column width="40%"}

```{r}
#| echo: false
#| warning: false
#| out-width: 100%

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(x = lifeexp)
  ) +
  geom_histogram(binwidth = 3)
```

:::

::::

---

## Histograms and binwidth: `binwidth = 10`

:::: columns

::: {.column width="60%"}

```{r}
#| eval: false

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(x = lifeexp)
  ) +
  geom_histogram(binwidth = 10)
```

:::

::: {.column width="40%"}

```{r}
#| echo: false
#| warning: false
#| out-width: 100%

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(x = lifeexp)
  ) +
  geom_histogram(binwidth = 10)
```

:::

::::

---

## Customizing histograms

:::: columns

::: {.column width="60%"}

```{r}
#| eval: false

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(x = lifeexp)
  ) +
  geom_histogram(binwidth = 3) +
  labs(
    x = "Female life expectancy (years)",
    y = "Frequency",
    title = "2023 Life Expectancy of Women"
  )
```

:::

::: {.column width="40%"}

```{r}
#| echo: false
#| warning: false
#| out-width: 100%

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(x = lifeexp)
  ) +
  geom_histogram(binwidth = 3) +
  labs(
    x = "Female life expectancy (years)",
    y = "Frequency",
    title = "2023 Life Expectancy of Women"
  )
```

:::

::::

---

## Fill with a categorical variable

:::: columns

::: {.column width="60%"}

```{r}
#| eval: false

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(
      x = lifeexp,
      fill = worldbankregion
    )
  ) +
  geom_histogram(
    binwidth = 3,
    alpha = 0.5
  ) +
  labs(
    x = "Female life expectancy (years)",
    y = "Frequency",
    title = "2023 Life Expectancy of Women"
  )
```

:::

::: {.column width="40%"}

```{r}
#| echo: false
#| warning: false
#| out-width: 100%

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(
      x = lifeexp,
      fill = worldbankregion
    )
  ) +
  geom_histogram(
    binwidth = 3,
    alpha = 0.5
  ) +
  labs(
    x = "Female life expectancy (years)",
    y = "Frequency",
    title = "2023 Life Expectancy of Women"
  )
```

:::

::::

---

## Facet with a categorical variable

:::: columns

::: {.column width="40%"}

<div style="font-size:0.80em;">

```{r}
#| eval: false

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(
      x = lifeexp,
      fill = worldbankregion
    )
  ) +
  geom_histogram(
    binwidth = 3,
    alpha = 0.5
  ) +
  labs(
    x = "Female life expectancy (years)",
    y = "Frequency",
    title = "2023 Life Expectancy of Women"
  ) +
  facet_wrap(~worldbankregion, nrow = 3)
```

</div>

:::

::: {.column width="60%"}

```{r}
#| echo: false
#| warning: false
#| out-width: 100%

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(
      x = lifeexp,
      fill = worldbankregion
    )
  ) +
  geom_histogram(
    binwidth = 3,
    alpha = 0.5
  ) +
  labs(
    x = "Female life expectancy (years)",
    y = "Frequency",
    title = "2023 Life Expectancy of Women"
  ) +
  facet_wrap(~worldbankregion, nrow = 3)
```

:::

::::

---

## Facet with a categorical variable (fixing labels)

:::: columns

::: {.column width="40%"}

<div style="font-size:0.80em;">

```{r}
#| eval: false

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(
      x = lifeexp,
      fill = worldbankregion
    )
  ) +
  geom_histogram(
    binwidth = 3,
    alpha = 0.5
  ) +
  labs(
    x = "Female life expectancy (years)",
    y = "Frequency",
    fill = "World Bank Region",
    title = "2023 Life Expectancy of Women"
  ) +
  facet_wrap(~worldbankregion, nrow = 3) +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank()
  )
```

</div>

:::

::: {.column width="60%"}

```{r}
#| echo: false
#| warning: false
#| out-width: 100%

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(
      x = lifeexp,
      fill = worldbankregion
    )
  ) +
  geom_histogram(
    binwidth = 3,
    alpha = 0.5
  ) +
  labs(
    x = "Female life expectancy (years)",
    y = "Frequency",
    fill = "World Bank Region",
    title = "2023 Life Expectancy of Women"
  ) +
  facet_wrap(~worldbankregion, nrow = 3) +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank()
  )
```

<div style="font-size:0.80em;">


Which version of all these plots seems most useful, and why?

</div>

:::

::::


---

# Density plot

---

## Density plot

:::: columns

::: {.column width="40%"}

<div style="font-size:0.80em;">

```{r}
#| eval: false

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(x = lifeexp)
  ) +
  geom_density()
```

A **density function** is a function whose value at any given point can be interpreted as providing a relative likelihood of values. That is, higher values of the density function indicate values of the random variable that are more likely to be observed.

</div>

:::

::: {.column width="60%"}



```{r}
#| echo: false
#| warning: false
#| out-width: 100%

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(x = lifeexp)
  ) +
  geom_density()
```

:::

::::

---

## Density plots: `adjust = 0.5`

:::: columns

::: {.column width="40%"}

<div style="font-size:0.80em;">

The *bandwidth* is a parameter that specifies the degree of smoothing. A small bandwidth shows many bumps (like a histogram with a small bin width), while a large bandwidth smooths over many bumps. It is specified using the *adjust* argument.

```{r}
#| eval: false

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(x = lifeexp)
  ) +
  geom_density(adjust = 0.5)
```

</div>

:::

::: {.column width="60%"}

```{r}
#| echo: false
#| warning: false
#| out-width: 100%

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(x = lifeexp)
  ) +
  geom_density(adjust = 0.5)
```

:::

::::

---

## Density plots: `adjust = 1`

:::: columns

::: {.column width="40%"}

<div style="font-size:0.80em;">

```{r}
#| eval: false

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(x = lifeexp)
  ) +
  geom_density(adjust = 1)
```

</div>

:::

::: {.column width="60%"}

```{r}
#| echo: false
#| warning: false
#| out-width: 100%

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(x = lifeexp)
  ) +
  geom_density(adjust = 1)
```

:::

::::

---

## Density plots: `adjust = 2`

:::: columns

::: {.column width="40%"}

<div style="font-size:0.80em;">

```{r}
#| eval: false

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(x = lifeexp)
  ) +
  geom_density(adjust = 2)
```

</div>

:::

::: {.column width="60%"}

```{r}
#| echo: false
#| warning: false
#| out-width: 100%

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(x = lifeexp)
  ) +
  geom_density(adjust = 2)
```

:::

::::

---

## Customizing density plots

:::: columns

::: {.column width="60%"}

<div style="font-size:0.80em;">

```{r}
#| eval: false

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(x = lifeexp)
  ) +
  geom_density(adjust = 1) +
  labs(
    x = "Female life expectancy (years)",
    y = "Density",
    title = "2023 Life Expectancy of Women"
  )
```

</div>

:::

::: {.column width="40%"}

```{r}
#| echo: false
#| warning: false
#| out-width: 100%

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(x = lifeexp)
  ) +
  geom_density(adjust = 1) +
  labs(
    x = "Female life expectancy (years)",
    y = "Density",
    title = "2023 Life Expectancy of Women"
  )
```

:::

::::

---

## Adding a categorical variable

:::: columns

::: {.column width="40%"}

<div style="font-size:0.80em;">

```{r}
#| eval: false

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(
      x = lifeexp,
      fill = worldbankregion
    )
  ) +
  geom_density(adjust = 1) +
  labs(
    x = "Female life expectancy (years)",
    y = "Density",
    title = "2023 Life Expectancy of Women",
    fill = "Region"
  )
```

</div>

:::

::: {.column width="60%"}

```{r}
#| echo: false
#| warning: false
#| out-width: 100%

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(
      x = lifeexp,
      fill = worldbankregion
    )
  ) +
  geom_density(adjust = 1) +
  labs(
    x = "Female life expectancy (years)",
    y = "Density",
    title = "2023 Life Expectancy of Women",
    fill = "Region"
  )
```

:::

::::

---

# Box plot

---

## Box plot

:::: columns

::: {.column width="40%"}

<div style="font-size:0.80em;">

```{r}
#| eval: false

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(x = lifeexp)
  ) +
  geom_boxplot() + theme(
  axis.title.y = element_blank(),
  axis.text.y  = element_blank(),
  axis.ticks.y = element_blank()
)
```

</div>

:::

::: {.column width="60%"}

```{r}
#| echo: false
#| warning: false
#| out-width: 100%

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(x = lifeexp)
  ) +
  geom_boxplot() + theme(
  axis.title.y = element_blank(),
  axis.text.y  = element_blank(),
  axis.ticks.y = element_blank()
)
```

:::

::::

---

## Box plot

:::: columns

::: {.column width="40%"}

```{r}
#| echo: false
#| warning: false
#| out-width: 100%

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(x = lifeexp)
  ) +
  geom_boxplot() + theme(
  axis.title.y = element_blank(),
  axis.text.y  = element_blank(),
  axis.ticks.y = element_blank()
)
```

:::

::: {.column width="60%"}

- Technical specifications vary across software packages
- **Median:** line in the middle of the box
- **Hinges** (25th and 75th percentiles): edges of the box
- **Upper whisker:** extends to the largest data point no more than \(1.5 \times IQR\) above the upper hinge (75th percentile); similarly for the lower whisker
- **Outliers:** observations beyond the whiskers, sometimes plotted 

:::

::::


---

## Comparing box plots and density plots

:::: columns

::: {.column width="60%"}

```{r}
#| echo: false
#| warning: false
#| fig-height: 6
#| out-width: 100%

library(patchwork)

p1 <-
  lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(aes(x = lifeexp)) +
  geom_density() +
  labs(
    x = "Female life expectancy (years)",
    y = "Density"
  )

p2 <-
  lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(aes(y = lifeexp)) +
  geom_boxplot() + theme(
  axis.title.y = element_blank(),
  axis.text.y  = element_blank(),
  axis.ticks.y = element_blank()
) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Female life expectancy (years)"
  )

p1 / p2
```

:::

::: {.column width="40%"}


- What features of the distribution do **both** plots show well?
- What additional information does the **density plot** provide?
- What additional information does the **box plot** provide?
- Is the distribution symmetric or skewed?

:::

::::

---

## Box plot

The distribution of population is highly skewed.

:::: columns

::: {.column width="50%"}

```{r}
#| echo: false
#| warning: false
#| out-width: 100%

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(aes(x = pop)) +
  geom_boxplot() +
  labs(x="Population")
```

:::

::: {.column width="50%"}

- Most countries have relatively small populations.
- A few countries have extremely large populations.
- These large populations appear as outliers.
- The long right tail indicates a **right-skewed** distribution.

:::

::::

---

## Box plot

<div style="font-size:0.85em;">

Here we plot the natural logarithm of the population because the original population values are highly skewed (primarily due to China and India). On the log scale, the largest countries are no longer extreme outliers, and the smallest populations become the outliers.

</div>

:::: columns

::: {.column width="40%"}

```{r}
#| echo: false
#| warning: false
#| out-width: 100%

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(aes(x = log(pop))) +
  geom_boxplot() + labs(x="Log(population)")
```

:::

::: {.column width="60%"}

<div style="font-size:0.85em;">

- The log (ln) transformation reduces the effect of extremely large populations.
- The distribution is much more symmetric than on the original scale.
- Countries with the smallest populations now appear as outliers.
- Log transformations are often useful for variables that span several orders of magnitude.

</div>

:::

::::

---

## Customizing box plots

:::: columns

::: {.column width="40%"}

<div style="font-size:0.80em;">

```{r}
#| eval: false

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(aes(x = lifeexp)) +
  geom_boxplot() +
  labs(
    x = "Female life expectancy (years)",
    y = NULL,
    title = "2023 Life Expectancy of Women"
  ) +
  theme(
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank()
  )
```

</div>

:::

::: {.column width="60%"}

```{r}
#| echo: false
#| warning: false
#| out-width: 100%

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(aes(x = lifeexp)) +
  geom_boxplot() +
  labs(
    x = "Female life expectancy (years)",
    y = NULL,
    title = "2023 Life Expectancy of Women"
  ) +
  theme(
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank()
  )
```

:::

::::

---

## Adding a categorical variable

:::: columns

::: {.column width="40%"}

<div style="font-size:0.80em;">

```{r}
#| eval: false

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(
      x = lifeexp,
      y = worldbankregion
    )
  ) +
  geom_boxplot() +
  labs(
    x = "Female life expectancy (years)",
    y = NULL,
    title = "2023 Life Expectancy of Women",
    subtitle = "By region"
  )
```

</div>

:::

::: {.column width="60%"}

```{r}
#| echo: false
#| warning: false
#| out-width: 100%

lifespan2023 %>%
  filter(sex == "Female") %>%
  ggplot(
    aes(
      x = lifeexp,
      y = worldbankregion
    )
  ) +
  geom_boxplot() +
  labs(
    x = "Female life expectancy (years)",
    y = NULL,
    title = "2023 Life Expectancy of Women",
    subtitle = "By region"
  )
```

:::

::::
